---
title: "Assignment 1"
output: html_document
author: Sofia Rodas
---

# Exploring Sea Level Rise as a topic in the New York Times

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load necessary libraries
library(jsonlite) # convert results from API queries into R-friendly formats 
library(tidyverse)
library(tidytext) # text data management and analysis
library(ggplot2) # plot word frequencies and publication dates
```

1.  Create a free New York Times account (<https://developer.nytimes.com/get-started>).

```{r API-key, include=FALSE}
# assign NYT Dev  API key

```


2.  Pick an environmental term of interest to you and use the {jsonlite} package to query the API. Pick something high profile enough and over a large enough time frame that your query yields enough articles for an interesting examination.

### API Function

Define the preconstructed function that let's us programatically call the API.

Note: The `Sys.sleep()` argument values were changed to trouble shoot issues with running the code on the desired phrase. 

```{r api-function}
nytAPI = function(term1, begin_date, end_date, api) {
  searchQ = URLencode(term1)
  url = paste0('https://api.nytimes.com/svc/search/v2/articlesearch.json?q=', searchQ,
               '&begin_date=', begin_date, '&end_date=', end_date, '&api-key=', api)
  
  initialsearch = fromJSON(url, flatten = TRUE) # we create an initial search to find out how many total hits, or articles that meet our query parameters
  hits = initialsearch$response$meta$hits
  
  if (hits == 0) return(data.frame()) ## if there are no hits, return an empty data frame

  maxPages = min(round(hits / 10) - 1, 10) # otherwise, we find out how many times we'll need to loop thru fcn, each time returning 10 pages

  df = data.frame(id=as.numeric(), created_time=character(),
                  snippet=character(), headline=character())

  for (i in 0:maxPages) {
    nytSearch = fromJSON(paste0(url, "&page=", i), flatten = TRUE)
    temp = data.frame(id = 1:nrow(nytSearch$response$docs),
                  created_time = nytSearch$response$docs$pub_date,
                  snippet = nytSearch$response$docs$snippet,
                  headline = nytSearch$response$docs$headline.main)
    df = rbind(df, temp)
    Sys.sleep(22)
  }
  return(df)
}
```

### Connect to the New York Times API and send a query.

```{r results}
# define the term of interest
term1 <- c("sea level rise")

# save the data to a csv
results <- nytAPI(term1 = term1, begin_date = "20240401", end_date = "20260401", api = API_KEY)

write.csv(results, paste0("NYT-", term1,".csv"))

# save the results from the API access for 
nytDat <- results
```


3.  Recreate the publications per day and word frequency plots using the snippets.

-   Make some (at least 3) transformations to the corpus (add context-specific stopword(s), simplify possessives, remove numbers)

Visualize the number of publications discussing sea level rise and the dates of each publication. 

```{r date-plot}
nytDat |> 
  mutate(pubDay = as.Date(created_time)) |> 
  group_by(pubDay) |> 
  summarise(count=n()) |> 
  filter(count >= 2) |> 
  ggplot() +
  geom_bar(aes(x=reorder(pubDay, count), y=count), stat="identity") +
  coord_flip() +
  theme_minimal() + 
  labs(title = "Count of New York Times articles on Sea Level Rise from April 2024-April 2026",
       y = "Count",
       x = "Publication date")
```

Climate week NYC took place from September 19, 2025 to September 29, 2025. September 19th, 2025 had the most articles about sea level rise. September 24th and 25th 2025 were also listed as having sea level rise appear twice.

### Tokenization of the snippets

The snippets are 
```{r plot_frequencies}
# save the snippets from the accessed data
snippet <-  names(nytDat)[4]

# use tidytext::unnest_tokens to put in tidy form
tokenized <- nytDat |> 
  unnest_tokens(word, snippet)
```
All of the filler words are still present. Let's remove the filler words.

### Stop Word Removal
```{r stop-words}

data(stop_words)
stop_words

 # remove stop words
tokenized <- tokenized |> 
  anti_join(stop_words)

#remove all numbers
clean_tokens <- str_remove_all(tokenized$word, "[:digit:]") 

# remove possessives
clean_tokens <- gsub("’s", '', clean_tokens) 

tokenized$clean <- clean_tokens

# drop where clean is empty
tokenized <- tokenized |> filter(clean != "")

# drop they're (for some reason this isn't being dropped with the stop words)
tokenized <- tokenized |> filter(clean != "they're")

# visualize the most used words
tokenized |> 
  count(clean, sort = TRUE) |> 
  filter(n > 5) |> 
  mutate(clean = reorder(clean, n)) |> 
  ggplot(aes(n, clean)) +
  geom_col() +
  labs(title = "Word frequencies in NYT article snippets about Sea Level Rise",
       y = NULL,
       x = "Count") + 
  theme_minimal()
  

```

4.  Recreate the publications per day and word frequency plots using the headlines variable (response.docs.headline.main). Compare the distributions of word frequencies between the snippet and headlines. Do you see any differences? Why might this be?

### Tokenization of the snippets

```{r plot_frequencies}

# save the snippets from the accessed data
# headlines <-  names(nytDat)[4]

# use tidytext::unnest_tokens to put in tidy form
headlines_tokenized <- nytDat |> 
  unnest_tokens(word, headline)
```


All of the filler words are still present. Let's remove the filler words.

Submit finished labs in both .Rmd and knitted PDF or html form to Canvas.

*\*\*Remember that complete assignments require readable, informative, aesthetically pleasing visualizations. You'll have to improve on the bare bones versions in this demo.*
