---
title: "local-Lab6:Word Embeddings"
author: "Mateo Robbins"
date: "2026-05-19"
output: html_document
---
```{r packages, include = FALSE}
library(tidytext)
library(tidyverse)
library(RSpectra)
library(broom)
library(textdata)
library(ggplot2)
library(dplyr)
```

In this lab we build word embeddings from scratch, then explore a set of pretrained GloVe embeddings trained on a much larger general-purpose text corpus. The goal is to understand how distributional context gets encoded as geometry — and what it means to do math on meaning.

We'll move through the core steps: computing co-occurrence statistics, weighting them with PMI, and reducing dimensionality with SVD to produce word vectors. Then we'll use those vectors to find semantically similar words and explore word arithmetic. We'll start off today by loading the climbing incident data again.

```{r data,}
incidents_df<-read_csv("https://raw.githubusercontent.com/MaRo406/EDS-231-text-sentiment/main/data/climbing_reports_model_dat.csv")
```

First, let's calculate the unigram probabilities -- how often we see each word in this corpus.

```{r unigrams}
unigram_probs <- incidents_df |> 
  unnest_tokens(word, Text) |> 
  anti_join(stop_words, by = "word") |> 
  count(word, sort = TRUE) |> 
  mutate(p = n/sum(n))
  
unigram_probs 
```

Now we have the probability of each word.

Next, we need to know how often we find each word near each other word -- the skipgram probabilities. In this case we'll define the word context as a five-word window. We'll slide that window across all of our text and record which words occurr together within that window.

Now let's write some code that adds an ngramID column that contains constituent information about each 5-gram we constructed by sliding our window.

```{r make-skipgrams}
skipgrams <- incidents_df |> 
  unnest_tokens(ngram, Text, token = "ngrams", n = 5) |> 
  mutate(ngramID = row_number()) |> 
  unite(skipgramID, ID, ngramID) |> 
  unnest_tokens(word, ngram) |> 
  anti_join(stop_words, by = "word")

skipgrams
```

Now we sum the total # of occurences of each pair of words.

```{r pairwise_count}
skipgram_probs <- 
```

The next step is to normalize these probabilities, that is, to calculate how often words occur together within a window, relative to their total occurrences in the data.

```{r norm-prob}
normalized_prob <- 

normalized_prob[5000:5010,]

#Which words are most associated with "rope"?   
normalized_prob 

```

Now we have all the pieces to calculate the point-wise mutual information (PMI) measure. It's the logarithm of the normalized probability of finding two words together. PMI tells us which words occur together more often than expected based on how often they occurred on their own.

Then we convert to a matrix so we can use matrix factorization and reduce the dimensionality of the data.

```{r pmi}
pmi_matrix<- normalized_prob %>%
    mutate(pmi = log10(p_together)) %>%
    cast_sparse(word1, word2, pmi)  ##tidytext::cast_sparse() to cast from tidy to a sparse matrix form from {Matrix}
```

Now we do the singular value decomposition with RSpectra::svds(). It's a "partial decomposition" as we are specifying a limited number of dimensions, in this case 100.

```{r pmi}
#remove missing data R Matrix package
pmi_matrix@x[is.na(pmi_matrix@x)] <- 0

pmi_svd <- svds(pmi_matrix, k = 100) #Reducing to 100 dimensions

#next we output the word vectors of left singular values.  
word_vectors <- pmi_svd$u
rownames(word_vectors) <- rownames(pmi_matrix)
```

These vectors in the "u" matrix are contain "left singular values". They are orthogonal vectors that create a 100-dimensional (k=100) semantic space where we can locate each word. The distance between words in this space gives an estimate of their semantic similarity.

Here's a function for matching the most similar vectors to a given vector.

```{r search-synonyms}
search_synonyms <- function(word_vectors, selected_vector, original_word = NULL) {
  similarities <- word_vectors %*% selected_vector / 
    (sqrt(rowSums(word_vectors^2)) * sqrt(sum(selected_vector^2)))
  
  result <- tibble(
    token      = rownames(word_vectors),
    similarity = as.numeric(similarities)
  ) %>%
    arrange(desc(similarity))
  
  if (!is.null(original_word)) result <- filter(result, token != original_word)
  result
}
```
Let's test it out!

```{r find-synonyms}

```

Here's a plot for visualizing the most similar words to a given target word.

```{r plot-synonyms}
slip %>%
    mutate(selected = "slip") %>%
    bind_rows(fall %>%
                  mutate(selected = "fall")) %>%
    group_by(selected) %>%
    top_n(15, similarity) %>%
    ungroup %>%
    mutate(token = reorder_within(token, similarity, selected)) %>%
    ggplot(aes(token, similarity, fill = selected)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~selected, scales = "free") +
    scale_x_reordered() +
    coord_flip() +
    theme(strip.text=element_text(hjust=0, size=12)) +
    scale_y_continuous(expand = c(0,0)) +
    labs(x = NULL, title = "What word vectors are most similar to slip or fall?")
  
```

One cool thing about representing words as numerical vectors is that we can use math on those numbers that has some semantic meaning.

```{r word-math}

```

So far we've built embeddings from scratch using a small, specialized corpus — climbing incident reports. Those embeddings capture meaning as it exists in that specific context. Now let's compare them to a set of pretrained embeddings trained on a much larger and more general corpus.
We'll be using GloVe (Global Vectors for Word Representation), a popular embedding method that builds co-occurrence statistics across a massive text corpus. The embeddings we're loading were trained on the Dolma dataset — a large open-source collection of web text, books, and other sources — giving us 300-dimensional vectors for over 1.2 million words.
Loading and cleaning the file takes a few steps. Each row is a word followed by 300 numbers, so we read it as a space-separated table, deduplicate any repeated tokens, and convert it into a matrix with words as row names — the same structure as our climbing embeddings.

```{r get-Glove}
glove_raw <- read.table("wiki_giga_2024_100_MFT20_vectors_seed_2024_alpha_0.75_eta_0.05.050_combined.txt", quote = NULL,sep = " ", header = F, fill = T)
```


```{r examine-GloVe}
gloVe <- tibble(glove_raw)
unique_glove <- distinct(gloVe)

# Should be a word, not a number
head(glove_raw$V1)

# Should be 301 (1 token + 300 dims)
ncol(glove_raw)
```

```{r GloVe-matrix}
# Deduplicate and build matrix
glove_vecs <- glove_raw[!duplicated(glove_raw$V1), ] %>%
  column_to_rownames(var = "V1") %>%
  as.matrix()
```

OK, that should do it.  Let's try some word math to test out these embeddings.

```{r GloVe-math}

```


### Assignment

#### Train Your Own Embeddings

1.  Using the data from your Nexis Uni query from the topic modeling lab, create a set of word embeddings. To do this, you'll essentially need to recreate the steps in today's example down through the chunk named "pmi".

2.  Think of 3-5 key words in your data set. Calculate and plot the 10 most semantically similar words for each of them. Identify and interpret any interesting or surprising results.

3.  Assemble 3 word math equations that you think could be useful or interesting for exploring the meaning of key words or ideas in your data set.

4.  Recreate the analyses in the last three chunks (find-synonyms, plot-synonyms, word-math) with the GloVe embeddings. How are they different from the embeddings created from the climbing accident data? Why do you think they are different?

#### Explore Pretrained Embeddings

5.  Recreate parts 2 and 3 above using the the GloVe embeddings in place of the ones you trained. How do they compare? What are the implications for applications of these embeddings?
