---
title: "Coral Bleaching Articles - Central America"
format: html
author: Sofia Rodas
---

```{r}
library(LexisNexisTools)
library(tidyverse)
library(quanteda)
library(tm)
library(topicmodels)
library(topicdoc)   
library(tidyverse)
library(tidytext)
library(reshape2)
library(tictoc)
library(LDAvis)
library(tsne)
library(servr)
```

```{r}

pre_files <- list.files(
  path = here::here("coral_bleaching_cen_amer", "data"),   # update to your folder
  pattern = "\\.docx$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

pre_dat <- lnt_read(pre_files)

pre_meta_df       <- pre_dat@meta
pre_articles_df   <- pre_dat@articles
pre_paragraphs_df <- pre_dat@paragraphs

df <- tibble(
  date     = pre_meta_df$Date,
  headline = pre_meta_df$Headline,
  id       = pre_meta_df$ID,
  text     = pre_articles_df$Article,
  source   = pre_meta_df$Newspaper
)
```

**1. Build and clean your corpus**
Create a corpus from your articles. Clean appropriately for your data —
consider what stopwords, minimum document frequencies, and tokenization
choices make sense given your specific corpus.

## Build the Corpus

We use `corpus()` from {quanteda} to build the corpus from our data
frame.

```{r corpus}
corpus <- corpus(x=df, textg_field = "text")
```

## Tokenize and Clean

`tokens()` handles punctuation and number removal. We also build a
custom stopword list that removes common journalistic filler words that
carry no thematic signal. We can return to this section to add
additional words once we see the results.

```{r tokens}
add_stops <- c(quanteda::stopwords("en", source = "smart"),
               "said", "says", "told", "according")

toks <- tokens(corpus, remove_punct = TRUE) |> 
  tokens_tolower()

toks1 <- tokens_select(toks, pattern = add_stops, selection = "remove")
```

## Build the Document-Feature Matrix

Topic modeling requires a document-feature matrix (DFM). We trim rare
terms (appearing in fewer than 2 documents) and remove empty rows — LDA
cannot process documents with no terms.

```{r dfm}
dfm1 <- dfm(toks1)
dfm2 <- dfm_trim(dfm1, min_docfreq = 2)

# Remove empty rows — LDA will error on these
sel_idx <- slam::row_sums(dfm2) > 0

dfm <- dfm2[sel_idx,]

```

------------------------------------------------------------------------

**2. Select k using multiple criteria**

Run three models (i.e. with 3 values of k) and select the overall best
value for k (the number of topics) - include some justification for your
selection: FindTopicsNumber() optimization metrics, theory,
interpretability, LDAvis. Select the best single value of k. Criteria to
consider:

\- **Optimization Metrics** - Run models across a range of k values and
produce a coherence plot using `topicdoc`.

\- **Theory**: what themes do you expect given your article topic?

\- **Interpretability**: read the top 10 terms at candidate k values —
do they form coherent, distinct themes?

\- **LDAvis**: do the topic circles separate well, or do they overlap
heavily?

Select a single best k and write 2–3 sentences justifying your choice
drawing on the three criteria.

## Step 1: Initial Model (Theory-Driven k)

Great, now we are almost ready to run a model. We just have to come up
with an initial value for k, the number of latent topics present in the
data. How do we do this? Let's say I think there may be political,
economic and environmental articles. So I will tell the model to look
for 3 topics by setting the k parameter = 3.

```{r LDA_k3}
k <- 3

topicModel_k3 <- LDA(dfm,
                     k,
                     method = "Gibbs",
                     control = list(iter = 100,
                                    verbose = 25,
                                    seed = 321))
```


### Inspect the Posterior Distributions

`posterior()` extracts the two key output matrices:

-   **theta (θ)**: document × topic matrix — *P(topic \| document)*
-   **beta/phi (φ)**: topic × term matrix — *P(term \| topic)*

```{r inspect_k3}
result <- posterior(topicModel_k3)

beta  <- result$terms
theta <- result$topics

cat("phi dimensions  (topics × vocab):", dim(beta), "\n")
cat("theta dimensions (docs × topics):", dim(theta), "\n")

# Top 10 terms per topic
terms(topicModel_k3, 10)

```


